Type Classes for Mathematics in Type Theory

### 2010-12-09 papers/mscs/mscs.tex
More referee-guided MSCS paper improvements.

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g.\ rewriting). Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully `unbundled' predicate-representations of algebraic structures practical to work with.

Since we intend to use this development as a basis for constructive analysis with practical certified exact real arithmetic, an additional objective and motivation in our design is to facilitate \emph{efficient} computation. In particular, we want to be able to effortlessly swap implementations of number representations. Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively. To express these interfaces elegantly and without duplication, our development\footnote{The sources are available
at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.

@@

The `native' notion of equality in Coq is that of terms being convertible, naturally reified as a proposition by the inductive type family \lstinline|eq| with single constructor \lstinline|eq_refl|:
\begin{lstlisting}
  eq_refl : ∀ (T: Type) (x: T), x ≡ x,
\end{lstlisting}
where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve the \lstinline|=| symbol for \emph{setoid} equality (to be discussed momentarily), as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions. We mention this explicitly only because such rewriting \emph{does} give rise to conditions when we depart from raw convertibility and introduce equivalence relations
that express how possibly distinct (unconvertible) terms may represent the same conceptual object. A typical example would be different formal integer fractions representing the same rational number. Rewriting a subterm using a proof of such an equality is permitted only if the subterm is argument to a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper} with respect to the equality in question, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

@@

The algebraic hierarchy of the \textsc{Ssreflect} libraries~\cite{Packed} uses an interesting (if fairly restrictive) alternative approach. It requires unicity of representation for all objects so that Leibniz equality is always applicable.

>>>>

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, equality of axiomatically posited and derived structure, idiomatic use of notations, support for models based on quotient representations, and convenient algebraic manipulation (e.g.\ rewriting). Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new solution based entirely on the use of Coq's new type class facility to make fully `unbundled' predicate-representations of algebraic structures practical to work with.

Our development is not merely aimed at formalization of theory, and our choice of a system based on type theory is no accident. It is our explicit ambition for the interfaces and theory we develop to be employed directly for the specification and parameterization of efficiently executable implementations of structures and procedures, expressed using type theory's native term reduction as a programming language. Thus, our work very much belongs in the long tradition of trying to realize to the greatest possible extent the promise of type theory to truly unite mathematics and certified (functional) programming, without making painful sacrifices on either side.

Because our ``ultimate'' goal is to use this development as a basis for constructive analysis with practical certified exact real arithmetic, and because numerical structures are ideal test subjects for our algebraic hierarchy, we shall use these to motivate and demonstrate the key parts of our development. Since we are concerned with \emph{efficient} computation, we want to be able to effortlessly swap implementations of number representations. Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively. To express these interfaces elegantly and without duplication, our development\footnote{The sources are available at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.

@@

The `native' notion of equality in Coq is that of term convertibility, naturally reified as a proposition by the inductive type family \lstinline|eq: ∀ (T: Type), T → T → Prop| with single constructor \lstinline|eq_refl|:
\begin{lstlisting}
  eq_refl : ∀ (T: Type) (x: T), x ≡ x,
\end{lstlisting}
where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve the `\lstinline|a = b|' notation for \emph{setoid} equality (to be discussed momentarily), as this is the equality we will be working with most of the time.

Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions. We mention this explicitly only because such rewriting \emph{does} give rise to conditions when we depart from raw convertibility and introduce equivalence relations that express how possibly distinct (unconvertible) terms may represent the same conceptual object. Rational numbers represented by (non-reduced) formal integer fractions are a typical example. Rewriting a subterm using a proof of such an equality is permitted only if the subterm is argument to a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper} with respect to the equality in question, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

@@

The algebraic hierarchy of the \textsc{Ssreflect} libraries~\cite{Packed} uses an alternative approach. It simply requires canonical representation of all objects, so that setoid equality is not needed. Of course, this policy severely restricts the freedom one has when implementing models of abstract structures. Indeed, for some sets, canonical representation schemes do not exist. The constructive reals, which are of particular interest to us, are an example of such a set.
